---
title: "GMM Tutorial"
author: "Katrina Jones & David Polly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This tutorial includes code and excersizes used in the 2018 Analytical Paleobiology Course, hosted at University of Florida. Lecture slide pdfs including theoretical content are provided at XXX

Code and datasets associated with this tutorial are available at https://github.com/katrinajones/qpal

#Intro to Geometric Morphometrics in R

##Part 1: Introduction to Geomorph Package

Geomorph provides a flexible tool for analyzing geometric morphometric data. For more information see `?(geomorph)`.

Loading up some data:

```{r, warning=F}
#load geomorph
library(geomorph, quietly=T)
library(qpal, quietly=T) #package including functions for the course

#Get example data - plethodon salamander heads
data("plethodon") #Load data into environment

#What's included in the dataset
summary(plethodon)

```

Converting between wide and long data formats:

```{r}
#Converting between wide and long formats
#Long to wide
wide<-two.d.array(plethodon$land)
wide[1:5,1:5]
#Wide to long
long<-arrayspecs(wide, 12, 2)
long[,,1]
```

Arrays are much more handy for handling GMM data in R because you can easily access all elements with simple code

```{r, results='hide'}
#Indexing arrays
#Get first specimen
long[,,1]
#Get first landmark of all specimens
long[1,,]
#Get X coordinates of all landmarks
long[,1,]
```

Geomorph data frames are a useful way of tidying all your data associated with your landmarks together into a single object, and are the native and sometimes required format for geomorph functions

```{r}
#Create geomorph data frame
plethgdf<-geomorph.data.frame(landmarks=long, species=plethodon$species, 
                              site=plethodon$site)
summary(plethgdf)
```

Once your data is all combined into a single geomorph data frame, its very useful to be able to subset it all at once, keeping track of all your separate variables. I’ve included a custom subsetting function I wrote for this purpose.

```{r, fig.show='hold'}
#plot first five specimens
plot(plethgdf$landmarks[,,1])

#plot raw data together
plotAllSpecimens(plethgdf$landmarks, mean=F)
```
```{r, results='hide'}
#plot in same shape space
Pcoords<-gpagen(plethgdf$landmarks)
plethgdf<-geomorph.data.frame(plethgdf, coords=Pcoords$coords, 
                              size=Pcoords$Csize)
```
```{r, fig.height=5, fig.width=5}
plotAllSpecimens(plethgdf$coords, mean=T, links=plethodon$links)

```

Comparing specimens

```{r, results='hide'}
#Are there any wierd specimens
plotOutliers(plethgdf$coords)
```
```{r, fig.show='hold'}
#Let's compare them
shape1<-mshape(plethgdf$coords)#consensus shape
shape2<-plethgdf$coords[,,14]

#Thin plate spline
plotRefToTarget(shape1, shape2, method=c("TPS"), mag=2)
#Lollipops
plotRefToTarget(shape1, shape2, method=c("vector"), mag=10)
#Points
plotRefToTarget(shape1, shape2, method=c("points"), mag=10)
```

##Part 2: Step-by-step morphometric analysis

This section provides a quick overview of the steps of a GMM analysis in R, which will be followed by a practical in which you will collect and run your own data.

The R package Stereomorph provides a nice interface for collecting landmarks and curves in 2D

```{r, eval=FALSE}
library(StereoMorph)
library(abind)

#Digitize 2D landmarks with Stereomorph
extdatafiles <- system.file("extdata", package="qpal")#find directory where external data are saved
setwd(extdatafiles)


#Name landmarks
lands<-c("condyle", "angular", "coronoid", "posterior molar", "tip incisor")
#Name curves
curves<-matrix(c("condtocor", "condtoang","condyle","condyle","coronoid",
                 "angular"), nrow=2,ncol=3)
#Digitize specimens
digitizeImage(image.file = 'ShrewsAndMarmots', shapes.file = 'teeth',
              landmarks.ref = lands, curves.ref = curves)
```
