---
title: "GMM Tutorial"
author: "Katrina Jones & David Polly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
extdatafiles <- system.file("extdata", package="qpal")
knitr::opts_knit$set(root.dir = extdatafiles)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
knitr::opts_chunk$set(fig.width=3, fig.height=3)
```


This tutorial includes code and excersizes used in the 2018 Analytical Paleobiology Course, hosted at University of Florida. Lecture slide pdfs including theoretical content are provided at XXX

Code and datasets associated with this tutorial are available at https://github.com/katrinajones/qpal

#Day 1 - Intro to Geometric Morphometrics in R

##Part 1: Introduction to Geomorph Package

Geomorph provides a flexible tool for analyzing geometric morphometric data. For more information see `?(geomorph)`.

Loading up some data:

```{r, warning=F}
#load geomorph
library(geomorph, quietly=T)
library(qpal, quietly=T) #package including functions for the course

#Get example data - plethodon salamander heads
data("plethodon") #Load data into environment

#What's included in the dataset
summary(plethodon)

```

Converting between wide and long data formats:

```{r}
#Converting between wide and long formats
#Long to wide
wide<-two.d.array(plethodon$land)
wide[1:5,1:5]
#Wide to long
long<-arrayspecs(wide, 12, 2)
long[,,1]
```

Arrays are much more handy for handling GMM data in R because you can easily access all elements with simple code

```{r, results='hide'}
#Indexing arrays
#Get first specimen
long[,,1]
#Get first landmark of all specimens
long[1,,]
#Get X coordinates of all landmarks
long[,1,]
```

Geomorph data frames are a useful way of tidying all your data associated with your landmarks together into a single object, and are the native and sometimes required format for geomorph functions

```{r}
#Create geomorph data frame
plethgdf<-geomorph.data.frame(landmarks=long, species=plethodon$species, 
                              site=plethodon$site)
summary(plethgdf)
```

Once your data is all combined into a single geomorph data frame, its very useful to be able to subset it all at once, keeping track of all your separate variables. I’ve included a custom subsetting function I wrote for this purpose.

```{r, fig.show='hold'}
#plot first five specimens
plot(plethgdf$landmarks[,,1])

#plot raw data together
plotAllSpecimens(plethgdf$landmarks, mean=F)
```
```{r, results='hide'}
#plot in same shape space
Pcoords<-gpagen(plethgdf$landmarks)
plethgdf<-geomorph.data.frame(plethgdf, coords=Pcoords$coords, 
                              size=Pcoords$Csize)
```
```{r, fig.height=5, fig.width=5}
plotAllSpecimens(plethgdf$coords, mean=T, links=plethodon$links)

```

Comparing specimens

```{r, results='hide'}
#Are there any wierd specimens
plotOutliers(plethgdf$coords)
```
```{r, fig.show='hold'}
#Let's compare them
shape1<-mshape(plethgdf$coords)#consensus shape
shape2<-plethgdf$coords[,,14]

#Thin plate spline
plotRefToTarget(shape1, shape2, method=c("TPS"), mag=2)
#Lollipops
plotRefToTarget(shape1, shape2, method=c("vector"), mag=10)
#Points
plotRefToTarget(shape1, shape2, method=c("points"), mag=10)
```

##Part 2: Step-by-step morphometric analysis

This section provides a quick overview of the steps of a GMM analysis in R, which will be followed by a practical in which you will collect and run your own data.

The R package Stereomorph provides a nice interface for collecting landmarks and curves in 2D. You must have either Chrome or Safari installed for this to work.

```{r, results='hide'}
library(StereoMorph)
library(abind)

#Digitize 2D landmarks with Stereomorph
extdatafiles <- system.file("extdata", package="qpal")#find directory where external data are saved
setwd(extdatafiles)
```
```{r, eval=FALSE}
#Name landmarks
lands<-c("condyle", "angular", "coronoid", "posterior molar", "tip incisor")
#Name curves
curves<-matrix(c("condtocor", "condtoang","condyle","condyle","coronoid",
                 "angular"), nrow=2,ncol=3)
#Digitize specimens
digitizeImage(image.file = 'ShrewsAndMarmots', shapes.file = 'teeth',
              landmarks.ref = lands, curves.ref = curves)
```

Read in the data

```{r, results='hide'}

#Look at data based on fixed landmarks only
teeth<-readShapes('teeth')
fixed<-teeth$landmarks.scaled

#Procrustes fit
gpa.lands <- gpagen(fixed)
```
```{r, fig.show='hold'}
plotAllSpecimens(fixed)
plot(gpa.lands)
```

Now add the curve sliders

```{r}
 #With Sliding semi-landmarks
curves<-teeth$curves.scaled

#resample curves to 3 sliders
curvelands<-list()
curvelands<-lapply(curves, function (x) lapply(x, pointsAtEvenSpacing, n=5))#Resample
curvelands<-lapply(curvelands, do.call, what=rbind)#merge
curvelands<-array(unlist(curvelands), dim=c(10,2,5))#convert to array
curvelands<-curvelands[c(2:4,7:9),,]#remove fixed points
rownames(curvelands)<-rep(c("cu1", "cu2"), each=3) #label
lands<-abind::abind(fixed, curvelands, along=1) #join with fixed
```
 
Need to make a 'sliders' file to tell geomorph which ones to slide during the procrustes fit. Here I'm using custom code which makes the sliders file based on the curve labels

```{r, results='hide'}
#make sliders file
sliders<-makesliders(rownames(lands),id=c("cu1", "cu2"), begin=c(1,1),
                     end=c(3,2))

#Procrustes fit
gpa.lands <- gpagen(lands, curves=sliders)
```
```{r}
plot(gpa.lands)
```

Now we'll try with a much bigger dataset, based on [insert citation]

```{r}
##Now lets try with a big dataset of shrewteeth!
#shrew teeth
data("shrewteeth")
summary(shrewteeth)
```
```{r, results='hide'}
#Procrustes fit
proc<-gpagen(shrewteeth)
```
```{r, fig.show='hold'}
#Plot raw data
plotAllSpecimens(shrewteeth)
#plot procrustes coordinates
plot(proc)
```

Now lets try putting it in a geomorph data frame

```{r, fig.height=5, fig.width=5}
#Make geomorph data frame
group<-as.factor(rep(c("a", "b"), length=dim(shrewteeth)[[3]]))#make up some labels
shrewdf<-geomorph.data.frame(raw=shrewteeth, coords=proc$coords,
                             size=proc$Csize, group=group)

#subsample by labels
a<-subsetgeom(shrewdf, "spec", which(shrewdf$group=="a"), keep=T)

#PCA
pca.lands <- plotTangentSpace(shrewdf$coords, label=TRUE,
                              groups = palette()[shrewdf$group])
shrewdf<-geomorph.data.frame(shrewdf, scores=pca.lands$pc.scores)
```

Now lets try an interactive 3D plot

```{r testgl, webgl=TRUE}
plot3d(shrewdf$scores[,1:3])#interactive 3D plots
text3d(shrewdf$scores[,1:3],texts=rownames(shrewdf$scores),pos=4,cex=.5)
```

Compare to the mean shape

```{r, fig.show='hold'}
#Consensus shape
consensus <- mshape(shrewdf$coords)
plotAllSpecimens(shrewdf$coords)
#centroid - midpoint of the landmarks
centroid <- apply(shrewdf$coords,2,mean)
points(centroid[1],centroid[2],col="Red", cex=2)

#Visualizing shapes
plotRefToTarget(consensus,shrewdf$coords[,,1])
```

Now try it out yourself! Take some photos of your friends faces, ideally from different angles, and try running your own morphometric analysis. Do different faces group separately? Or does error associated with photo angle swamp out the signal?

#Day 2 - By-hand GMM analysis and statistical framework

##Procrustes superimposition

Calculating Procrustes distances by hand

```{r}
#Calculate distance
A<-shrewdf$raw[,,1]
B<-shrewdf$raw[,,2]
pdist<-sqrt(sum((A-B)^2))
pdist

#Consensus shape
consensus <- mshape(shrewdf$coords)

#Procrustes distance
dists <- array(dim=dim(shrewdf$coords)[3])
for(i in 1:dim(proc$coords)[3]) 
{ dists[i] <- sqrt(sum((shrewdf$coords[i]-consensus)^2)) }
head(sort(dists))
```

Calculate centroid size

```{r}
#Centroid size
shape <- A
centroid <- apply(shape,2,mean)
centroidsize <- sqrt(sum((t(t(shape)-centroid))^2))

plot(shape, xlim=c(-800,800), ylim=c(-800,800))
points(t(centroid), col="red", pch=19)
```

Run through the steps of a Procrustes Superimposition: Translation, scaling, rotation

Translation

```{r, fig.show='hold'}
#Translation
newshape<-t(t(shape)-centroid)

#plot
plot(shape, xlim=c(-800,800), ylim=c(-800,800), main="Translate")
points(newshape, col="blue")
points(t(centroid), col="red")
newcent<-apply(newshape, 2, mean)
points(t(newcent), pch=19,col="Red")

#centroid size doesn't change
newcentsize<-sqrt(sum((t(t(newshape)-newcent))^2))
centroid
newcentsize

#scale
resizedshape <- newshape / centroidsize

#plot
plot(c(-1,1),c(-1,1),xlab="x",ylab="y", type="n", main="Scale and Rotate")
points(t(newcent), pch=19)
points(resizedshape, pch=19,col="Red")

angle <- 45 * pi / 180
rotmat <- matrix(c(cos(angle),-sin(angle),sin(angle),cos(angle)),byrow=T,ncol=2)

#plot
points(resizedshape%*%rotmat, pch=19,col="Blue")
```

##Principal Components Analysis

Three methods for calculating total variance which should always give the same results

```{r}
nspec<-dim(shrewdf$raw)[3]
nspec
sum(apply(shrewdf$coords, 3, function (x) sum((x-consensus)^2)))/(nspec-1)
sum(pca.lands$sdev^2)
sum(pca.lands$pc.scores^2)/(nspec-1)
```

PCA Calculation

```{r}
#convert procrustes coordinates to wide format for use outside geomorph
coords2d<-two.d.array(shrewdf$coords) 
consensusvec<-apply(coords2d, 2, mean)#consensus in vector format

#calculate residuals
resids<-sweep(proc$coords, c(1,2), consensus)

#calculate covariance matrix
P<-cov(two.d.array(resids))

#single value decomposition
pca.stuff<-svd(P)
eigenvalues <- pca.stuff$d
eigenvectors <- pca.stuff$u
scores <- two.d.array(resids)%*%eigenvectors
```

Calculate variances again to check

```{r}
sum(apply(coords2d,2,var)) # total variance of Procrustes coordinates
sum(apply(two.d.array(resids),2,var)) # total variance of Procrustes residuals
sum(pca.stuff$d) # total variance of singular values
sum(apply(scores, 2, var)) # total variance of scores
```

Compare plots

```{r, fig.show='hold'}
pca<-plotTangentSpace(shrewdf$coords, warpgrids=F)
plot(scores[,1:2])
```

##Statistical analysis of GMM data

